---
title: "Vitiligo age-of-onset and PRS Association"
author: "Genevieve HL Roberts"
date: "2023-12-31"
output:
  pdf_document:
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'asis', message = FALSE, warning = FALSE)
```

```{r packages, include=FALSE}
library(tidyverse)
library(broom)
library(here)
library(mixtools)
library(gridExtra)
library(pandoc)
library(pander)

source(here("functions.R"))
```

# Introduction

In this study, we aim to investigate the association between a vitiligo 
Polygenic Risk Score (PRS) excluding MHC Class II region SNPs and the 
age-of-onset. Our hypothesis posits a correlation wherein a higher PRS 
(excluding MHC Class II) is linked to an earlier age-of-onset, 
indicating a potential association with elevated genetic risk.

Moreover, we extend our hypothesis to carriers of the extreme risk/early age 
of onset MHC Class II haplotype. In this subgroup, we anticipate that the 
impact of a high PRS without MHC Class II may be diminished due to the 
substantial genetic risk conferred by the haplotype (haplotype OR=8.1). 
We believe there may be an interaction effect, suggesting that the slope 
of the regression line for age-of-onset ~ PRS is smaller in individuals 
carrying the high-risk haplotype compared to non-carriers.


```{r readin}
# Read genotypes data
risk_genos <- read.table(here("sensitive_data/GWAS123_forRiskScore_inR.csv"), sep = ",", header = TRUE) %>%
  mutate(vitiligo = ifelse(PHENOTYPE==2, 1, 0)) %>%
  mutate(sex = ifelse(SEX==2, 1, 0))  %>%
  mutate(IID = toupper(IID)) %>%
  mutate(eAOO_hap_carrier =
           case_when(
             RS145954018 >= 1 & RS114448410 >=1 ~ 1L,
             RS145954018 >= 1 ~ NA_integer_,
             TRUE ~ 0))

# Read in PCs
pcs <- read.table(here("sensitive_data/GWAS123_DanielIDs_Phenos_PCs.txt"), header=TRUE) %>%
  select(IID, starts_with("g")) %>%
  mutate(IID = toupper(IID))

kCovariates <- pcs %>%
  select(starts_with("g")) %>%
  names()
kCovariates <- c(kCovariates, "sex")

#read in AOO
aoo <- read.table(here("sensitive_data/GWAS123G3RepageonsetonlyFinal_from_Ying_AOO.csv"),
                  header=TRUE, sep=",") %>%
  mutate(IID = toupper(newID)) %>%
  select(IID, VITageonset)


# Read risk score weights
weights <- read.table(here("Risk_Score_Weights_no_TagSNPs.csv"),
                      sep = ",", header = TRUE) %>%
  #removes the MHC Class II SNPs, which has special coding
  filter(!SNP %in% c("RS114448410", "RS145954018:RS114448410")) 
```

```{r recode alleles}
# Recode SNPs to make sure everything is w/r/t risk allele
snps_to_recode <- weights %>%
  filter(Recode. == "Yes") %>%
  pull(SNP)

risk_genos_recoded <- recode_snps(df=risk_genos,
                                  snps_to_recode=snps_to_recode)
```

```{r compute PRS}
# compute the main PRS (does not include MHC Class II)
main_prs_no_mhcII <- calculate_main_prs(risk_genos = risk_genos_recoded,
                           weights = weights,
                           iid_col = "IID",
                           weights_col = "Weights_risk_direction_only") %>%
  rename(no_mhc_classII_prs = score)

#compute MHC only PRS
mhcII_only_prs <- calculate_class2_risk_score(risk_genos = risk_genos_recoded)

#join with the main dataframe and with AOO data
analysis_df <- risk_genos %>%
  left_join(main_prs_no_mhcII) %>%
  left_join(mhcII_only_prs) %>%
  left_join(pcs) %>%
  left_join(aoo) %>%
  mutate(CONFIRMED_prs = no_mhc_classII_prs + mhc_class2_only_prs)

#select the columns that are prs
prs_cols <- analysis_df %>%
  select(ends_with("prs")) %>%
  names()

#scale the PRS with respect to the mean and sd in controls
analysis_df <- scale_prs(data = analysis_df,
                         pheno_col = "vitiligo",
                         pheno_cond = 0,
                         scale_columns = prs_cols)

#make a case only analysis df with non-missing age-of-onset
case_only_analysis_df <- analysis_df %>%
  filter(!is.na(VITageonset))
```

\newpage
# Results

## Categorizing age-of-onset into Early- and Late- Onset Groups
I no longer have the AOO classifications that we created previously, so I 
have re-fit the finite mixture model to categorize GWAS123 cases as either
early onset or late onset.  I have used the same methods here.

In the eAOO paper it says,
"After subgroup assignment and control matching, 
the combined early-onset subgroup contained 704 cases and 9,031 controls 
and the combined late-onset subgroup contained 1,467 cases and 19,156 controls."

```{r fmm to get aoo groups}
# fit a finite mixture model with 2 groups to get early and late
# onset group designations
set.seed(123)
fmm_mat <- as.matrix(case_only_analysis_df$VITageonset)
rownames(fmm_mat) <- case_only_analysis_df$IID

fit_fmm <- normalmixEM(fmm_mat, k = 2)
eaoo_posterior <- bind_cols(case_only_analysis_df$IID,
                            fit_fmm$posterior[,1],
                            fit_fmm$posterior[,2]) 
colnames(eaoo_posterior) <- c("IID",
                              "eAOO_posterior_probability",
                              "lAOO_posterior_probability")

#categorize as "eAOO" if posterior probability of eAOO group is >=80%
eaoo_posterior <- eaoo_posterior %>%
  mutate(AOO_category = 
           case_when(
             eAOO_posterior_probability >= 0.8 ~ "early_onset",
             lAOO_posterior_probability >= 0.8 ~ "late_onset",
             TRUE ~ NA_character_
           ))


analysis_df <- analysis_df %>%
  left_join(eaoo_posterior) %>%
  mutate(early_onset_vitiligo =
           case_when(
             AOO_category == "early_onset" & vitiligo == 1 ~ 1L,
             vitiligo == 0 ~ 0L,
             TRUE ~ NA_integer_
           ),
         late_onset_vitiligo =
           case_when(
             AOO_category == "late_onset" & vitiligo == 1 ~ 1L,
             vitiligo == 0 ~ 0L,
             TRUE ~ NA_integer_
           ),
         eA00_rs145954018 = RS145954018,
         generic_rs9271597 = RS114448410
         )

case_only_analysis_df <- analysis_df %>%
  filter(!is.na(VITageonset))

summarize_aoo_cat <- case_only_analysis_df %>%
  group_by(AOO_category) %>%
  summarize(n = n(),
            `Actual Median` = median(VITageonset),
            `Actual Mean` = mean(VITageonset),
            `Actual Minimum` = min(VITageonset),
            `Actual Maximum` = max(VITageonset),
            `Actual SD` = sd(VITageonset)
  ) %>%
  mutate(`FMM Mean` = ifelse(AOO_category == "early_onset", fit_fmm$mu[1],
                           ifelse(AOO_category == "late_onset", fit_fmm$mu[2], NA)),
         `FMM SD` = ifelse(AOO_category == "early_onset", fit_fmm$sigma[1],
                           ifelse(AOO_category == "late_onset", fit_fmm$sigma[2], NA))) %>%
  transmute(
    `Age of Onset Category` = AOO_category,
    n,
    `FMM Mean`,
    `Actual Mean`,
    `FMM SD`,
    `Actual SD`,
    `Actual Minimum`,
    `Actual Maximum`
  )

pandoc.table(summarize_aoo_cat,
             round = 2,
             caption="Early and Late Onset Subgroup Descriptive Statistics")
```

As shown in the table above, you can see that we achieve similar sample sizes
for early and late onset (719 early here vs. 704 in our earlier analysis
and 1,431 late here vs. 1,467 in our earlier analysis).The FMM-estimated 
distribution means are also identical, at 10.3 here
versus 10.3 for GWAS123 + Rep in our earlier analysis and 34.0
here versus 34.0 for GWAS123 + Rep in our earlier analysis.

\newpage
```{r}
dist_params  <- summarize_aoo_cat %>%
  filter(!is.na(`FMM Mean`) & !is.na(`FMM SD`)) %>%
  transmute(AOO_category = `Age of Onset Category`,
            FMM_mean = `FMM Mean`,
            FMM_sd = `FMM SD` )

histo_plot <- plot_distributions(histogram_data = case_only_analysis_df,
                               histogram_variable = "VITageonset",
                               dist_df = dist_params,
                               name_col = "AOO_category",
                               mean_col = "FMM_mean",
                               sd_col = "FMM_sd",
                               size = dim(case_only_analysis_df)[1],
                               binwidth = 2)
histo_plot
```

\newpage
## Associations with Vitiligo Risk (Case-Control)

Next, I wanted to verify that the associations looked as I would expect
for each of the PRSs. So, I checked the association with overall vitiligo
risk (i.e. in all cases and controls in GWAS123):

```{r}
#test the association of each PRS with vitiligo risk
scaled_prs_cols <- analysis_df %>%
  select(ends_with("_prs_scaled")) %>%
  names()

phenotypes <- c("vitiligo", "early_onset_vitiligo", "late_onset_vitiligo")
predictors <- c(scaled_prs_cols, "eA00_rs145954018", "generic_rs9271597", "eAOO_hap_carrier")

combos <- crossing(phenotypes, predictors) %>% arrange(desc(phenotypes))

results_df <- pmap_df(combos,
                      ~ logistic_regression(
                        data = analysis_df,
                        covariates = kCovariates,
                        pheno = ..1,
                        prs = ..2)) %>%
  filter(term == PRS) %>%
  select(-term, -estimate, -std_error) %>%
  mutate(pval = formatC(pval, format = "e", digits = 2)) %>%
  mutate(PRS = str_remove(PRS, "_prs_scaled"))  %>%
  rename(`PRS or SNP` = PRS)

pandoc.table(results_df, round=2)
```

These results look generally as I would expect, with the CONFIRMED risk score
having the highest performance with respect to OR per standard deviation of PRS.
The P-values = 0 mean that the P-value is lower than the numerical precision in R,
which I believe is something like P < 1e-300.

Note that the scaling is different for SNPs and for PRS, so the estimates
are not directly comparable (for PRS, the OR is per standard deviation of PRS, 
whereas for SNPs, it is per SNP risk allele).

\newpage
## Associations with Vitiligo Age-of-Onset (Case Only)

I also checked the association with vitiligo age-of-onset
risk (i.e. cases with non-missing age-of-onset in GWAS123):
```{r}
#test the association with vitiligo age-of-onset
aoo_results_df <- pmap_df(list(prs = predictors),
                      ~ aoo_linear_regression(
                        data = case_only_analysis_df,
                        covariates = kCovariates,
                        pheno = "VITageonset",
                        prs = ..1)) %>%
  filter(term == PRS) %>%
  select(-term, - std_error) %>%
  mutate(pval = formatC(pval, format = "e", digits = 2)) %>%
  mutate(PRS = str_remove(PRS, "_prs_scaled")) %>%
  rename(`PRS or SNP` = PRS)

pandoc.table(aoo_results_df, round=2)
```

Again, these results look about as I would expect: the MHC Class II only PRS
is more strongly associated than the confirmed PRS. Nevertheless, there
is modest association between the non-MHC Class II PRS and lower vitiligo
age-of-onset, as we hypothesized (P=7.30e-04).

Note that the scaling is different for SNPs and for PRS, so the estimates
are not directly comparable (for PRS, the estimate is interpreted as years
per standard deviation of PRS, whereas for SNPs, it is interpreted
as years per copy of SNP risk allele).

```{r}
ggplot(analysis_df, aes(y = VITageonset,
                        x = no_mhc_classII_prs_scaled)) +
  geom_point(alpha = 0.3, size = 0.4, aes(color = factor(AOO_category))) + 
  geom_smooth(method = "lm", se = TRUE, color="black") +
  labs(title = "Correlation Between Vitiligo Age-of-Onset and PRS",
       x = "Vitiligo PRS (MHC Class II Excluded)",
       y = "Vitiligo Age-of-Onset",
       color = "Age-of-Onset Category") +
    theme_bw()
```

\newpage
## Interaction Test Between High-Risk MHC Haplotype and PRS (excluding MHC Class II SNPs)

My hypothesis was that the vitiligo PRS becomes less relevant when combined with
the eAOO high-risk haplotype — essentially, that the genetic risk is already
fairly saturated for individuals that carry the high-risk MHC Class II haplotype.

As a result, I expected to observe a correlation slope close to zero for
age-of-onset ~ PRS in those *with* the eAOO haplotype, indicating a lack of
association. Conversely, for those *without* the high-risk haplotype,
I anticipated a correlation slope for age-of-onset ~ PRS significantly
less than 1.

The interaction test, which compares slopes between carriers and non-carriers,
serves as a formal assessment of this hypothesis.
Notably, achieving statistical significance in an interaction test
may have inadequate power, as a sample size of ~16X the sample size to detect
the main effect is typically needed to detect an interaction effect.

```{r}
#test the interaction
aoo_int_res_df <- pmap_df(list(pheno = c("VITageonset")),
                      ~ regression_interaction(
                        data = case_only_analysis_df,
                        interactor = "eAOO_hap_carrier",
                        covariates = kCovariates,
                        predictor= "no_mhc_classII_prs_scaled",
                        outcome = ..1)) %>%
  filter(str_detect(term, ":")) %>%
  mutate(term = str_remove(term, "_scaled"))

pandoc.table(aoo_int_res_df, round=2)
```

```{r}
ggplot(analysis_df, aes(y = VITageonset,
                        x = no_mhc_classII_prs_scaled,
                        color = factor(eAOO_hap_carrier))) +
  geom_point(alpha = 0.3, size=0.4) +  # Make points slightly transparent
  geom_smooth(method = "lm", se = TRUE, aes(group = eAOO_hap_carrier)) +  # Fitted regression line without confidence interval
  labs(title = "Correlation between age-of-onset and PRS,\nStratified by High Risk Haplotype Carrier",
       x = "Vitiligo PRS (MHC Class II Excluded)",
       y = "Vitiligo Age-of-Onset",
       color = "Early Onset High Risk Haplotype Carrier") +
  scale_color_manual(values = c("0" = "#66c2a5", "1" = "#fc8d62"),
                     labels = c("0" = "Non-Carrier", "1" = "Carrier")) +
  theme_bw()
```

Although there is a trend towards a stronger negative slope for those that do
not carry the high-risk eAOO haplotype relative to those that do carry
the haplotype (consistent with our hypothesis), the difference 
is not significant (interaction P = 0.38).

### Stratified Association in Early and Late Onset Groups
Due to power limitations in the interaction test, I performed a stratified
analysis in the early- and late-onset groups separately.

```{r}
#test the association with vitiligo age-of-onset
eAOO_res <- aoo_linear_regression(
                        data = filter(case_only_analysis_df, early_onset_vitiligo==1),
                        covariates = kCovariates,
                        pheno = "VITageonset",
                        prs = "no_mhc_classII_prs_scaled") %>%
  filter(term == PRS) %>%
  mutate(AOO_category = "early_onset")

lAOO_res <- aoo_linear_regression(
                        data = filter(case_only_analysis_df, late_onset_vitiligo==1),
                        covariates = kCovariates,
                        pheno = "VITageonset",
                        prs = "no_mhc_classII_prs_scaled") %>%
  filter(term == PRS) %>%
  mutate(AOO_category = "late_onset")

strat_aoo_results_df <- bind_rows(eAOO_res, lAOO_res) %>%
  select(-term, - std_error) %>%
  mutate(pval = formatC(pval, format = "e", digits = 2)) %>%
  mutate(PRS = str_remove(PRS, "_prs_scaled")) %>%
  rename(`PRS` = PRS) %>%
  select(AOO_category, everything())

pandoc.table(strat_aoo_results_df,
             round=2,
             caption = "Stratified Association between PRS and Age-of-Onset in Early- and Late-Onset Groups")
```

```{r}

categorized_only_analysis_df <- case_only_analysis_df %>%
  filter(!is.na(AOO_category))

ggplot(categorized_only_analysis_df, aes(y = VITageonset,
                        x = no_mhc_classII_prs_scaled)) +
  geom_point(alpha = 0.3, size = 0.4, aes(color = factor(AOO_category))) + 
  geom_smooth(method = "lm", se = TRUE, color="black") +
  facet_grid(~AOO_category) + 
  labs(title = "Correlation Between Vitiligo Age-of-Onset and PRS
       Stratified by Early- and Late- Onset Groups",
       x = "Vitiligo PRS (MHC Class II Excluded)",
       y = "Vitiligo Age-of-Onset",
       color = "Age-of-Onset Category") +
    theme_bw() +
  geom_text(data = strat_aoo_results_df, aes(label = paste0("P-value=", pval)),
             x = Inf, y = 85, hjust = 1, vjust = 1, size = 3, color = "black")
```



